### 2012-01-26 docs/notes.tex
reorg and precision plot

Methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions. One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between blocks.   Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.

@@

In general, we may have different types of events that can occur.   In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.  As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.  % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} 
% \end{align}
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \} 
% \end{align}
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
\end{align}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.  Note the statistics $\mathbf{s}_{ij}$ also do not change until an event involving either $i$ or $j$ occurs.  We model the intensity functions via a log linear model
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z}).
\end{align}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.  This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.  Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

@@

We construct $\mathbf{s}_{ij}$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}. Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{ijm}} = \tau_{ijm}$). If $(i,j)$ has never occurred, we set $v_{ijm}=0$.  
\begin{enumerate}
\item Sender out-degree: $s_{ij1}(t) = \sum_{m:t_m<t} I(i_m=i)$
\item Sender in-degree: $s_{ij2}(t) = \sum_{m:t_m<t} I(j_m=i)$
\item Receiver out-degree: $s_{ij3}(t) = \sum_{m:t_m<t} I(i_m=j)$
\item Receiver in-degree: $s_{ij4}(t) = \sum_{m:t_m<t} I(j_m=j)$
\item Reciprocity (AB-BA): $s_{ij5}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-taking (AB-BY): $s_{ij6}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XA): $s_{ij7}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XB): $s_{ij8}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-continuing (AB-AY): $s_{ij9}(t) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Number of shared two-paths: $s_{ij10}(t) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% \item Shared contacts (since last occurrence): $s_{ij10}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% \item Triadic closure (since last occurrence): $s_{ij11}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% \item Shared contacts (since last occurrence): $s_{ij12}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
\end{enumerate}

@@

\subsection{Extensions}
TODO: Explain difficulties in doing a mixed membership model.  As in the mixed-membership stochastic blockmodel \cite{Airoldi2008}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.  In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.  One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.

TODO: Discuss other variants, e.g. HMM that guides which latent class over time?

TODO: What about different clusters for senders and receivers (cite our KDD paper)?
\section{Inference}

@@

\section{Simulation}
\begin{figure}
\center
%\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
%\includegraphics[width=3in]{../figs/syn-counts.pdf}
%\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2. Right: Loglikelihood vs. iteration during MCMC.}
\label{fig:syncounts}
\end{figure}

@@

\subsection*{Prediction experiment}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the posterior mean given the training data.  Recall at a given cutoff $k$ is found by first sorting the predicted intensity functions at each time, then computing the mean rank of the observed events in that list.

>>>>

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}. One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between blocks.   Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.

@@

It is straightforward extend this to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ of possible events include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.  As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.  % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} 
% \end{align}
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \} 
% \end{align}
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
\end{align}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.  Note the statistics $\mathbf{s}_{ij}$ also do not change until an event involving either $i$ or $j$ occurs.  We model the intensity functions via a log linear model
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
\end{align}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.  This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.  Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

@@

We construct $\mathbf{s}_{ij}$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}. Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{ijm}} = \tau_{ijm}$). If $(i,j)$ has never occurred, we set $v_{ijm}=0$.  
\begin{enumerate}
\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i)$
\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i)$
\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)$
\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)$
\item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
\end{enumerate}

@@

\subsection{Extensions}
TODO: Explain difficulties in doing a mixed membership model.  As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.  In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.  One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.

TODO: Discuss other variants, e.g. HMM that guides which latent class over time?  Or a DP instead of fixed K?

TODO: What about different clusters for senders and receivers (cite our KDD paper)?

TODO: What about doing a similar thing with topic model statistics in s(t,i,j)?

\section{Inference}

@@

\section{Simulation}
\begin{figure}
\center
\includegraphics[width=1.5in]{../figs/syn/mat.pdf}
\includegraphics[width=2.5in]{../figs/syn/counts.pdf}
\includegraphics[width=2in]{../figs/syn/bias.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2. Right: Loglikelihood vs. iteration during MCMC.}
\label{fig:syncounts}
\end{figure}

@@

\subsection*{Prediction experiment}
\begin{figure}
\center
\includegraphics[width=3.5in]{../figs/syn/test-recall.pdf}
\caption{Predictive performance on test data.  Left: Recall on synthetic example shown in Figure \ref{fig:syncounts}.}
\label{fig:recall}
\end{figure}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the posterior mean given the training data.  Recall at a given cutoff $k$ is found by first sorting the predicted intensity functions at each time, then computing the mean rank of the observed events in that list.
